---
title: "Some lecture"
output: html_notebook
---

#FactoMineR package

```{r}
  library(FactoMineR)
  data(decathlon)
```

```{r}
  tail(decathlon)
```
```{r}
str(decathlon)
```

```{r}
myPCA=PCA(decathlon, quanti.sup=11:12, quali.sup=13)
```
```{r}
myPCA
```

```{r}
plot(myPCA, habillage = 13)
```
#karpov and Lorenzo - far separeted.
#the values of variables are very diffent, e.g.
#to prove it:
```{r}
decathlon["Uldal", ] #entie row
```


```{r}
decathlon["Karpov", ]
```
#for Discus e.g., Javeline, not big diffenece but still:
#as these too point are most different, others are even more close
#comparison close guys
```{r}
decathlon["Karpov", ]
```


```{r}
decathlon["Clay", ]
```
#geometrically, they are close, and values closed as well.
```{r}
myPCA2=PCA(decathlon, quali.sup=13)
```

```{r}
myPCA2
```
#2dims plot, analyzing original valriables - columns;
#
```{r}
plot(myPCA2, habillage = 13)
```
#more variables - all, 1 to 12 instead of 11-12

```{r}
myPCA2$ind$coord
```

#first 5 are major dimenstions
```{r}
PC1scores=myPCA2$ind$coord[, 1]
PC1scores
```

```{r}
y100=decathlon[,1]
y100
```

```{r}
plot(as.numeric(PC1scores), y100)
```

```{r}
cor(as.numeric(PC1scores), y100) #correlation is -0.7081629
```

```{r}
myPCA$ind$cos2
```


```{r}
X=read.csv("https://web.stanford.edu/~hastie/ElemStatLearn/datasets/zip.digits/train.2",header = FALSE) # read from the on-line source
tail(X,1) # list the last line
```


```{r}
m3=matrix(as.numeric(X[17,]),16,16) # take image 39 (line of matrix X) and present it as a 16x16 matrix
image(-m3[,ncol(m3):1], axes = F, col = grey(seq(0, 1, length = 256))) # command image plots the greyscale image of 2
```



```{r}
PCA(X)
```

```{r}
m3=matrix(as.numeric(X[671,]),16,16) # take image 39 (line of matrix X) and present it as a 16x16 matrix
image(-m3[,ncol(m3):1], axes = F, col = grey(seq(0, 1, length = 256))) # command image plots the greyscale image of 2
```

```{r}
m3=matrix(as.numeric(X[212,]),16,16) # take image 39 (line of matrix X) and present it as a 16x16 matrix
image(-m3[,ncol(m3):1], axes = F, col = grey(seq(0, 1, length = 256))) # command image plots the greyscale image of 2
```
```{r}
m3=matrix(as.numeric(X[565,]),16,16) # take image 39 (line of matrix X) and present it as a 16x16 matrix
image(-m3[,ncol(m3):1], axes = F, col = grey(seq(0, 1, length = 256))) # command image plots the greyscale image of 2
```

```{r}
m3=matrix(as.numeric(X[601,]),16,16) # take image 39 (line of matrix X) and present it as a 16x16 matrix
image(-m3[,ncol(m3):1], axes = F, col = grey(seq(0, 1, length = 256))) # command image plots the greyscale image of 2
```

