---
title: "Home assignment 02: Data analysis Task 4"
output: html_notebook
---

This is the second Almaz Murzabekov's home assignment of Data analysis task: 4

### 4. Assignment on Correspondence Analysis

Tarvel Review Ratings Dataset: https://archive.ics.uci.edu/ml/datasets/Tarvel+Review+Ratings

This data set is populated by capturing user ratings from Google reviews. Reviews on attractions from 9 categories across Europe are considered. Google user rating ranges from 1 to 5 and average user rating per category is calculated.
  
- Attribute 1 : Unique user id
- Attribute 2 : Average ratings on churches
- Attribute 3 : Average ratings on resorts
- Attribute 4 : Average ratings on beaches
- Attribute 5 : Average ratings on parks
- Attribute 6 : Average ratings on theatres
- Attribute 7 : Average ratings on museums
  
  
2. Use FactoMineR package to:
```{r}
data = 
  as.data.frame(
    read.table("data/t4/google_review_ratings.csv", header = TRUE, sep = ",")
  )

head(data)
```


##### (a) Do the χ2 test for independence and interpret it, see Section 2.2.2 of [4].
```{r}
chisq.test(data[, 2:7])
```
Chi-square test results shows us that we cannot reject the null hypothesis. Null hypothesis tells us that there are not any statistically dependence between components of dataset.

##### (b) Perform the CA, get the 2D representation of row and column profiles See p.87 of [4]

```{r}
ca_result = CA(data[, 2:7], graph = FALSE)
plot(ca_result, invisible = "col", title = "CA results, ROWS")
plot(ca_result, invisible = "row", title = "CA results, COLS")
plot(ca_result)
```

##### (c) Analyze the patterns obtained in item 2b. Focus on the total variability, similarities/dissimilarities and the conclusions that can be made from the simultaneous representation of rows and columns. See examples, [4], pp. 92-125.

```{r}
ca_result
```

##### (d) Provide the table and graph of eigenvalues, justify the choice of principal components.
```{r}
round(ca_result$eig, 2)
```

Eigenvalues correspond to the amount of information retained by each axis. Dimensions are ordered decreasingly and listed according to the amount of variance explained in the solution. Dimension 1 explains the most variance in the solution, followed by dimension 2 and so on.

The cumulative percentage explained is obtained by adding the successive proportions of variation explained to obtain the running total. For instance, 55.73% plus 35.61% equals 91.24%, and so forth. Therefore, about 91% of the variation is explained by the first two dimensions.
    
In our analysis, the first two axes explain 88.6% of the variation. This is an acceptably large percentage.

```{r}
barplot(ca_result$eig[, 1], names.arg = paste("dim", 1 : nrow(ca_result$eig)))
```
    
##### (e) Discuss the quality of the CA representation based on cos2 for rows and columns, [4], p.87.
```{r}
ca_result$col$cos2
```
The result of the CA shows that the the second column (Category. 2) well described by dimension 1 with accuracy 0.99%, but the same dimention doesnot reprecent Category 3 as well. So, the sum of occuracies of categories is slightly equal to one.
    
```{r}
ca_result$row$cos2
```
The result of the CA shows that the two dimensions 1 and 2 are sufficient to retain 91% of the total variation contained in the rows
If a row item is well represented by two dimensions, the sum of the cos2 is closed to one. For example, row1 is presented in CA with 99% occuracy (0.43 + 0.56), but also, we have examples where the accuracy extremelly low (row 25 - 35% occuracy)


### REFERENCES
[1] Dataset: https://archive.ics.uci.edu/ml/datasets/Tarvel+Review+Ratings#
[2] https://www.displayr.com/use-not-use-correspondence-analysis/
